Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-08-08 TexSubmission/answer_referees.tex

{\em Secondly in the algorithm presented on page 9, it says to draw $M$ samples from $\zeta_k$ while not mentioning
what to do with these. This should be specified in the manuscript.}

\vspace{.3cm}

The way the unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$ was computed was indeed not clear. In (7) and at the bottom of page 6 we only stated the existence of a random variable $\zeta_k$ such that $\widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta_k)$ is an unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$.

At each time step, we then sample independently $(\zeta^m_k)_{1\le m \le M}$ and define the following estimator of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$:
\[
\frac{1}{M}\sum_{m=1}^M \widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta^m_k)\eqsp.
\]
This is now detailed clearly just after (7).

{\em Maybe if the time continuous process was denoted by $\widetilde{X}$ or something similar the difference would be clear.}

\textcolor{red}{Il est gentil Michel mais introduire une notation pour ca moi ca ne me plait pas. On repond au reste, je suis d'avis de lui dire merci mais non merci.}

{\em
On page 3, you talk about not being able to extend the FFBSi algorithm to this class of models, but
as far as I understand it the same sampling technique should be possible to draw from the approximative
smoothing distribution.
}

The usual FFBSi cannot be extended easily for the same reasons as the FFBS (it relies on the approximation of the backward kernel which is a ratio as defined in (6) in our paper). But it is true that the linear version of the FFBSi may be extended in a similar way as PaRIS algorithm (while it does stil require a backward pass). The introduction was modified to make this statement clear.

@@

{\em On page 15, line 13, I believe that $h(m)$ has sneaked its way into the equation and should be removed.}

\vspace{.3cm}
 $h(m)$ was removed in this equation.

\subsection*{Referee 2}

\subsubsection*{Major comments}
\begin{enumerate}
\item
\item {\em When presenting equation (2), the authors should explain the importance of approximating those type of functionals.}

The importance of these quantities is now clearly stated after equation (2). We added the usual applications to the computation of the Fisher score and the intermediate quantity of the EM algorithm. We also highlighted that it was crucial for online inference of latent data models which supports the use of the proposed online smoother.

\item {\em General practical strategies to specify $\vartheta_k$, $p_k$, $\hat{\sigma}_+^k$ in the SDE context should be
discussed.}

Choices that can be made for  $\vartheta_k$ and $p_{k-1}$ are discussed after (4) and the definition of $\phi_k^N$. The specific approximation of the fully adapted particle filter is also illustrated in the numerical section. \textcolor{red}{Pour le sigma Pierre tu as peut-etre plus d'elements. On peut bien sur parler des conditions de l'EA mais je pense qu'une intuition pratique est plus percutante ici.}

\item {\em Conditions (A1)-(A3) should be carefully discussed. How restrictive are they? For
example, what if $\phi(\omega_s)$ (or $L_\omega$) is not bounded below? Is there any general solution
in this case?}

\item
\item
\item
\item
\item {\em Can anything be said about the bias of the proposed estimator for finite $N$? Is it
only due to the form of  $\Lambda_k^N$ ?}

\item {\em What can be said about a CLT for the proposed estimator?}
\textcolor{red}{On peut le faire, pas tres dur, mais ca va allonger les choses et sans message tres clair sur la variance asymptotique a mon avis. Voila une suggestion de reponse...}

A CLT was derived for the original version of PaRIS algorithm as the number $N$ of particles grows to infinity. A CLT may also derived for our extension as it is a noisy version of the original method. However, even if its proof is not too involved it still require additional assumptions and technicalities. And the asymptotic variance remains hard to analyze (for instance to choose the parameters in comparison to other smoothing methods). Analyzing the variance of particle smoothers (FFBS, FFBSi, PaRIS, two-filer) is at the core of ongoing works and we believe that comparing the variance would be more suitable and interesting than just providing a CLT with no rules to choose between methods. We are also working on the estimation of such asymptotic variances which is a complex problem out of the scope of this paper.

\item {\em It is worth mentioning that, for the log-growth model, $X$ is a function of $\sigma$ and what
are the consequences of that in a context where $\sigma$ is an unknown parameter.}

\item {\em I quite like the choice of $Q(\theta;\theta)$ in the examples. It would be very interesting to
have an implementation of the actual EM algorithm for parameter estimation in the
examples provided, using GRand Paris on the E step.}

The numerical section now contains a complete implementation of the EM algorithm which highlights clearly \textcolor{red}{BIM} the performance of the proposed PaRIS algorithm.  
\end{enumerate}

\subsubsection*{Minor comments}
\begin{enumerate}
\item{\em page 3, line 36: The authors should clearly state that the proposed methodology
does not rely on time discretisation approximations.}

This is now explicitly written in the introduction.

\item{\em page 5, line 21: computing $\longrightarrow$ computed.}

This typo was corrected.

\item {\em page 8, line 52: I suggest "... $L_w$ is bounded below almost surely, ..."}

This was changed as suggested.

\item {\em page 11, line 10: does the fixed lag algorithm have the same cost for all lag values?
What is the computational time of one run of GRand PaRIS for the examples presented?}

\textcolor{red}{M. Gloaguen, your call !}

\item {\em page 16, line 49: For the proof to be valid, the induction result needs to be established for $k\ge 0$.}

\end{enumerate}

### 2017-08-08 TexSubmission/ParisEM.tex
First answers to referee 2

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.

In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles. The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.

@@

Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$. Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
\begin{equation}
\label{eq:importance:weights}
\omega^{\ell}_k := \frac{\qk(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_k^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. Its crucial property is that it does not require a backward pass, the smoothed expectation is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.

PaRIS algorithm relies on the following fundamental property of $T_k[H_k]$ when $H_k$ is as in \eqref{def:addfunc}:
\begin{align*}
T_k[H_k](X_k) &=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k)\middle|X_k,Y_{0:k-1} \right]\eqsp,\\
&= \frac{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)\left\{T_{k-1}[H_{k-1}](x_{k-1}) + h_{k-1}(x_{k-1},X_k)\right\}}{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)}\eqsp.
\end{align*}
Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$ (starting with $\tau^i_0 = 0$, $1\le i\le N$), approximating $T_k[H_k](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$. First, replacing $\phi_{k-1}$ by $\phi^N_{k-1}$ in the last equation leads to the following approximation of $T_k[H_k](\xi^i_k)$:
\begin{equation}
\label{eq:Tk}
T_k^N[H_k](\xi_k^i) = \sum_{j=1}^N \Lambda_{k-1}^N(i,j)\left\{T_{k-1}[H_{k-1}](\xi_{k-1}^j) + h_{k-1}(\xi^j_{k-1},\xi^i_k)\right\}\eqsp, 
\end{equation}
where
\begin{equation}
\label{eq:Lambda}
\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
\end{equation}
Computing exactly these approximations would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}. Therefore, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
\begin{enumerate}[(i)]
\item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by \eqref{eq:Lambda}.
\item \label{it:PaRIS:smooth} Set
\[
\tau^{i}_{k+1} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_{k}^{i,\ell}}_{k} + h_{k} \left(\xi^{J_{k}^{i,\ell}}_{k}, \xi^{i}_{k+1}\right)  \right\}\eqsp.
\]
\end{enumerate}
Then, \eqref{def:addfunc} is approximated by
\[
\phi_{0:n\vert n}^N[\tau_n] = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
\] 
As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$. As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept-reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.

@@

\begin{figure}[p]
\centering
%\includegraphics[scale = .5]{obs_SINE}
\caption{{\em SINE model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
\label{fig:res:SINE:obs}
\end{figure}

>>>>

The exact computation of these expectations is usually not possible in the case of partially observed diffusions. In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
 \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights. Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly. Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}. Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein. However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}. To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).  In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.

 The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}. 
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}. 
Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions. Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals. 
Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.  One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements. The total number $n$ of observations is fixed in this paper to describe the algorithm but it could process any additional observation straightforwardly.

In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities. The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates. The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights. The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother. The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.

@@

Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
\[
\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
\begin{equation}
\label{def:addfunc}
\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
\end{equation}
when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models. These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}). They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.

The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
\begin{equation}
\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
\end{equation}
The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.

At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$. 
For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_{k-1}$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$. Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computed following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_{k-1}(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
\begin{equation}
\label{eq:importance:weights}
\omega^{\ell}_k := \frac{q_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\end{equation}
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_k^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
The most simple choice for $p_{k-1}$ and $\vartheta_k$ is the bootstrap filter proposed by \cite{gordon:salmond:smith:1993} which sets $p_{k-1} = q_{k-1}$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = 1$. In the case of partially observed SDE $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made. The approximation can be obtained using a discretization scheme such as Euler method or a Poisson based approximation as detailed below. A more appealing choice is the fully adapted particle filter which sets for all $x,x'\in\mathbb{R}^d$, $p_{k-1}(x,x') \propto q_{k-1}(x,x')g_k(x')$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = \int q_{k-1}(x,x')g_k(x')\mu(\rmd x')$. Here again $q_{k-1}$ has to be replaced by an approximation. In Section~\ref{sec:exp}, it is replaced by the Gaussian approximation provided by a Euler scheme which leads to a Gaussian proposal density $p_{k-1}$ as the observation model is linear and Gaussian.

PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. Its crucial property is that it does not require a backward pass, the smoothed expectation is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.

PaRIS algorithm relies on the following fundamental property of $T_k[H_k]$ when $H_k$ is as in \eqref{def:addfunc}:
\begin{align*}
T_k[H_k](X_k) &=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k)\middle|X_k,Y_{0:k-1} \right]\eqsp,\\
&= \frac{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)\left\{T_{k-1}[H_{k-1}](x_{k-1}) + h_{k-1}(x_{k-1},X_k)\right\}}{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)}\eqsp.
\end{align*}
Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$ (starting with $\tau^i_0 = 0$, $1\le i\le N$), approximating $T_k[H_k](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$. First, replacing $\phi_{k-1}$ by $\phi^N_{k-1}$ in the last equation leads to the following approximation of $T_k[H_k](\xi^i_k)$:
\begin{equation}
\label{eq:Tk}
T_k^N[H_k](\xi_k^i) = \sum_{j=1}^N \Lambda_{k-1}^N(i,j)\left\{T_{k-1}[H_{k-1}](\xi_{k-1}^j) + h_{k-1}(\xi^j_{k-1},\xi^i_k)\right\}\eqsp, 
\end{equation}
where
\begin{equation}
\label{eq:Lambda}
\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
\end{equation}
Computing exactly these approximations would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}. Therefore, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
\begin{enumerate}[(i)]
\item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by \eqref{eq:Lambda}.
\item \label{it:PaRIS:smooth} Set
\[
\tau^{i}_{k+1} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_{k}^{i,\ell}}_{k} + h_{k} \left(\xi^{J_{k}^{i,\ell}}_{k}, \xi^{i}_{k+1}\right)  \right\}\eqsp.
\]
\end{enumerate}
Then, \eqref{def:addfunc} is approximated by
\[
\phi_{0:n\vert n}^N[\tau_n] = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
\] 
It is clear from steps (i) to (iii) that each time a new observation $Y_{n+1}$ is received, the quantities $(\tau_{n+1}^i)_{1\le i \le N}$ can be updated only using $Y_{n+1}$, $(\tau_n^i)_{1\le i \le N}$ and the particle filter at time $n$. This means that storage requirements do not increase when processing additional data.

As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$. As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept-reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.

@@

\begin{figure}[p]
\centering
\includegraphics[scale = .5]{obs_SINE}
\caption{{\em SINE model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
\label{fig:res:SINE:obs}
\end{figure}
